On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-05-19 Manuscript/manuscript.tex
Went through some of Prof Schifanos revisions

%3)
%The International Association of Athletics Federations (also referred to
%as IAAF or World Athletics), which is the governing board for international running
%competitions, commissioned a study in 2009 regarding the reaction time 
%barrier and whether the threshold should be lowered. 
%The paper analyzed sprinters reaction times and the biological processes 
%that went into reacting to the starting gun and concluded 
%that athletes can react in as low as 0.08 seconds.  
%One study by Pain looked at
Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
\eds{are hurdlers generally considered a subset of sprinters or are they 
considered a separate type of athlete?  Specifically did the study above 
recommend the barrier be lowered for both sprinters and hurdlers?
You also refer to sprinters elsewhere in the manuscript.  Are those instances
meant to refer to both sprinters and hurdlers, as well?}
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
\eds{is this true? are you still using the 2019 world championships data for the
rank-based analyses, or only the national competitions?}
\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
international and 2019 international vs 2022 international.  Should I instead say
that there are three types of data across 2 sections or something like that?}
\eds{I think this is fine here, but it is not at all clear that there are two
subsections in the rank-based results for 2022 national vs 2022
international and 2019 international vs 2022 international.}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
\eds{With the following newly added sentences, I think we can leave this paragraph here.}
We can also see from Figure~\ref{fig:Boxplots} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period. 
%Something that is extremely important to note in the data is the inconsistency 
%of World Athletics in terms of how they penalize false starts.  
From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Championships) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}. 
%This was desirable for World Athletics as false starts can make already long 
%track meets more tedious, both for athletes and viewers watching the television 
%broadcast.  
%\eds{I think we lost something here with all of the edits so I removed the 
%previous sentence. What was desirable for World Athletics -- the 2007-2009 rule
% change?  Why was this desirable?  The rule change would seem to make the event
% longer and not shorter? } 
There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female false starts at the 2009 World 
Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced men's 
false starts by two thirds in 2011 (6 male and 4 female disqualifications) 
\citep{iaaf2009falsestart}.
%The effect? At the 2011 
%World Championships there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under more lenient rules.

@@

This dataset is too small to perform any meaningful statistical analysis, but
the data were expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).  \eds{this is where you can put the relevant 
information about the other competitions that was commented out in the 
introduction.}
\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\of{Is the one below what you're talking about?}
\eds{Yes, this is the part from the intro I was referring to, but you need to 
edit this paragraph now so it all works together.}
First we can compare the reaction times of athletes who
competed at the World Track and Field Championships in both 2019 and 2022, and 
second we can compare how athletes performed at the 2022 championships relative 
to their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.

>>>>

%3)
%The International Association of Athletics Federations (also referred to
%as IAAF or World Athletics), which is the governing board for international running
%competitions, commissioned a study in 2009 regarding the reaction time 
%barrier and whether the threshold should be lowered. 
%The paper analyzed sprinters reaction times and the biological processes 
%that went into reacting to the starting gun and concluded 
%that athletes can react in as low as 0.08 seconds.  
%One study by Pain looked at
Reaction time in sprint and hurdle events has been studied by many authors.
\citet{pain2007sprint} examined
reaction times for nine male athletes.  The
fastest athlete of the nine had, on average (even with their two fastest times 
removed), a reaction time of 0.087 seconds and a standard deviation of 0.004
seconds.  Two other athletes were able to react similarly when under certain
muscular tightening conditions \citep{pain2007sprint}. 
In a study commissioned by International Association of Athletics Federations 
(IAAF), \citet{komi2009iaaf} recommended that the disqualification 
barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
\eds{are hurdlers generally considered a subset of sprinters or are they 
considered a separate type of athlete?  Specifically did the study above 
recommend the barrier be lowered for both sprinters and hurdlers?
You also refer to sprinters elsewhere in the manuscript.  Are those instances
meant to refer to both sprinters and hurdlers, as well?}
\of{Yes I would say that hurdlers are a subset of sprinter.  The reaction time
barrier for the men's 100 meters and men's 110 meter hurdle are both 0.1 seconds.
The paper recommended for the reaction time barrier to be decreased to roughly 0.8
seconds.  It does not reference any distinction between hurdlers/sprinters, so I
think they can be grouped together.  In my personal opinion I don't see why the
reaction time for hurdlers versus 100m dash would differ significantly.  In terms
of what I wrote in the paper, I want to stay accurate to what is in the study
but also generalize studies involving reaction times to include hurdlers.  I believe
many of the studies examining the reaction time barrier from a physiological
perspective to find the minimum possible reaction time do not deal with hurdles.
Thus, I don't know if I can change what I wrote to be both accurate and understandable,
but what do you both think?}
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body
movement. This could provide a more accurate representation of reaction time, 
and thus the possibility of a false disqualification would be lowered.
The possibility of using a video-based medium has been proved to be a reliable
measure of reaction time \citep{mudric2015evaluation}.

@@

There are two types of data that are explored throughout this paper: data from
1999-2022 for every world championship, and data from athletes who competed in
the 2022 World Championships and another competition (either 2019 World 
Championships or a 2022 national competition typically held within three months 
of the 2022 World Championships). 
\eds{is this true? are you still using the 2019 world championships data for the
rank-based analyses, or only the national competitions?}
\of{Yes the ranked based comparison has 2 subsections: 2022 national vs 2022
international and 2019 international vs 2022 international.  Should I instead say
that there are three types of data across 2 sections or something like that?}
\eds{I think this is fine here, but it is not at all clear that there are two
subsections in the rank-based results for 2022 national vs 2022
international and 2019 international vs 2022 international.}
\of{Currently under the Methods/Results section I have GLMM and rank based comparison
do you think I need subsections under rank based comparison, or is that too much
compartmentalization?  Or should I instead say there are 3 types of data that are
used in two different methods, or something like that?}
The first data set is used for the GLMM analyses described in 
Section~\ref{sec:glmm}, while the second collection of data is used for the 
rank-based comparisons described in Section~\ref{sec:rank}.

@@

\eds{I wonder if this paragraph would be better placed in the introduction?}
\of{I think it could fit in the brief section between the start of the Data section
and the GLMM subsection in the paragraph starting with "There are two types..."}
\eds{With the following newly added sentences, I think we can leave this paragraph here.}
\of{I agree I think this works well.}
We can also see from Figure~\ref{fig:Boxplots} that the reaction times clearly
differ from year to year. The venue of the World Championships changes each 
year, so weather or climate related factors such as humidity, precipitation, 
elevation, may indeed play a role, but additionally, both the technology and 
false start penalization rules have changed over the study period. 
%Something that is extremely important to note in the data is the inconsistency 
%of World Athletics in terms of how they penalize false starts.  
From 2007--2009,
the IAAF (the former name for World Athletics, and the governing body for the 
World Track and Field Championships) instituted a rule change that allowed one 
warning false start before a sprinter was disqualified \citep{iaaf2009falsestart}. 
%This was desirable for World Athletics as false starts can make already long 
%track meets more tedious, both for athletes and viewers watching the television 
%broadcast.  
%\eds{I think we lost something here with all of the edits so I removed the 
%previous sentence. What was desirable for World Athletics -- the 2007-2009 rule
% change?  Why was this desirable?  The rule change would seem to make the event
% longer and not shorter? } 
There were 18 male and 7 female false starts at the 2007 World 
Championships, and 18 male and 7 female false starts at the 2009 World 
Championships. 
Starting in 2011, the rule was scrapped, and the old policy which automatically 
disqualifies runners who false start was reinstated. By returning to the harsher 
policy and cracking down on false starts, World Athletics had reduced men's 
false starts by two thirds in 2011 (6 male and 4 female disqualifications) 
\citep{iaaf2009falsestart}.
%The effect? At the 2011 
%World Championships there were 6 male disqualifications and 4 female 
%disqualifications \citep{iaaf2009falsestart}.  
%One study that looked at different false start rules that IAAF has
%imposed from 1997 to 2009 and found that the reaction times when there was less 
%leniency improved by a statistically significant amount
%\citep{haugen2013effect}. 
\citet{haugen2013effect} examined the effect of different false start
rules that IAAF has imposed from 1997 to 2009 and found statistically 
significant improvement in reaction times under more lenient rules.

@@

\eds{this is where you can put the relevant 
information about the other competitions that was commented out in the 
introduction.}
\eds{Can you show some descriptive statistics for this data?  Total number of 
clusters? Average number of international vs national competitions across 
athletes/clusters?}
\of{Is the one below what you're talking about?}
\eds{Yes, this is the part from the intro I was referring to, but you need to 
edit this paragraph now so it all works together.}
The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete). First we can compare the reaction times of athletes who
competed at the World Track and Field Championships in both 2019 and 2022, and 
then we can compare how athletes performed at the 2022 championships relative 
to their national/qualifiers meet. In order to reduce any chances
of athletes improving reaction time, we can also look at data from qualifying meets.
Almost every athlete had to qualify by proving they are elite at the national level, 
usually by coming in at least fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August. The equipment used to measure reaction
time cannot be guaranteed to be the same from national to international
competition, but the time span between meets has been reduced considerably.
